---
title: "Assertive R Programming with assertr"
author: "Tony Fischetti"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Assertive R Programming with assertr}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


In data analysis workflows that depend on un-sanitized data sets from external
sources, it’s very common that errors in data bring an analysis to a
screeching halt. Oftentimes, these errors occur late in the analysis and
provide no clear indication of which datum caused the error.

On occasion, the error resulting from bad data won’t even appear to be a
data error at all. Still worse, errors in data will pass through analysis
without error, remain undetected, and produce inaccurate results.

The solution to the problem is to provide as much information as you can about
how you expect the data to look up front so that any deviation from this
expectation can be dealt with immediately. This is what the assertr package
tries to make dead simple.

Essentially, assertr provides a suite of functions designed to verify
assumptions about data early in an analysis pipeline. assertr is meant to be
used with the piping constructs of the `magrittr` package and fits right in with
the structure and data manipulation verbs of the `dplyr` package.

### concrete data errors

Let’s say, for example, that the R’s built-in car dataset, mtcars, was not 
built-in but rather procured from an external source that was known for making
errors in data entry or coding.

In particular, the mtcars dataset looks like this:
```{r}
head(mtcars)
```

But let's pretend that the data we got accidentally negated the 5th mpg value:

```{r}
our.data <- mtcars
our.data$mpg[5] <- our.data$mpg[5] * -1
our.data[4:6,]
```

Whoops!

If we wanted to find the average miles per gallon for each number of engine
cylinders, we might do so like this:

```{r message=FALSE}
library(dplyr)

our.data %>%
  group_by(cyl) %>%
  summarise(avg.mpg=mean(mpg))

```

This indicates that the average miles per gallon for a 8 cylinder car is a lowly
12.43. However, in the correct dataset it's really just over 15. Data errors
like that are extremely easy to miss because it doesn't cause an error, and the
results look reasonable.

### enter assertr

To combat this, we might want to use assertr's `verify` function to make sure
that `mpg` is a positive number:

```{r error=TRUE, purl = FALSE}
library(assertr)

our.data %>%
  verify(mpg >= 0) %>%
  group_by(cyl) %>%
  summarise(avg.mpg=mean(mpg))
```

If we had done this, we would have caught this data error.

The `verify` function takes a data frame (its first argument is provided by
the `%>%` operator), and a logical (boolean) expression. Then, `verify`
evaluates that expression using the scope of the provided data frame. If any
of the logical values of the expression's result are `FALSE`, `verify` will
raise an error that terminates any further processing of the pipeline.

We could have also written this assertion using `assertr`'s `assert` function...

```{r error=TRUE, purl = FALSE}
our.data %>%
  assert(within_bounds(0,Inf), mpg) %>%
  group_by(cyl) %>%
  summarise(avg.mpg=mean(mpg))
```

The `assert` function takes a data frame, a predicate function, and an arbitrary
number of columns to apply the predicate function to. The predicate function
(a function that returns a logical/boolean value) is then applied to every
element of the columns selected, and will raise an error when it finds the
first violation.

Internally, the `assert` function uses `dplyr`'s `select` function to extract
the columns to test the predicate function on. This allows for complex
assertions. Let's say we wanted to make sure that all values in the dataset
are *greater* than zero (except `mpg`):

```{r error=TRUE, purl = FALSE}
library(assertr)

our.data %>%
  assert(within_bounds(0,Inf, include.lower=FALSE), -mpg) %>%
  group_by(cyl) %>%
  summarise(avg.mpg=mean(mpg))
```

### verify vs. assert

The first noticable difference between `verify` and `assert` is that `verify`
takes an expression, and `assert` takes a predicate and columns to apply it to.
This might make the `verify` function look more elegant--but there's an
important drawback. 

`verify` has to evaluate the entire expression first, and *then* check if there
were any violations. Because of this, `verify` can't tell you the offending
datum. That brings us to the second difference.

Because `assert` applies the predicate function to each datum, one at a time,
it can stop immediately after finding the first violation, and specify the
location and the value of the offending element.

This also means that `assert` will fail sooner than `verify`, potentially
making it a faster, less time-consuming affair for data that are assumed to
have errors. 

One important drawback to `assert`, and a consequence of its application of
the predicate to *columns*, is that `assert` can't confirm assertions about
the data structure *itself*. For example, let's say we were reading a dataset
from disk that we know has more than 100 observations; we could write a check
of that assumption like this:

```{r eval=FALSE, purl = FALSE}
dat <- read.csv("a-data-file.csv") %>%
  verify(nrow(dat) > 100) %>%
  ....
```

This is a powerful advantage over `assert`... but `assert` has one more
advantage of its own that we heretofore ignored.

### assertr's predicates

`assertr`'s predicates, both built-in and custom, make `assert` very powerful.
The three predicates that are built in to `assertr` are

- `not_na` - that checks if an element is not NA
- `within_bounds` - that returns a predicate function that checks if a numeric
value falls within the bounds supplied, and
- `in_set` - that returns a predicate function that checks if an element is
a member of the set supplied.

We've already seen `within_bounds` in action... let's use the `in_set` function
to make sure that there are only 0s and 1s (automatic and manual, respectively)
values in the `am` column...


```{r, eval=FALSE, purl = FALSE}
our.data %>%
  assert(in_set(0,1), am) %>%
  ...
```

If we were reading a dataset that contained a column representing boroughs of
New York City (named `BORO`), we can verify that there are no mis-spelled
or otherwise unexpected boroughs like so...

```{r, eval=FALSE, purl = FALSE}
boroughs <- c("Bronx", "Manhattan", "Queens", "Brooklyn", "Staten Island")

read.csv("a-dataset.csv") %>%
  assert(in_set(boroughs), BORO) %>%
  ...
```

Rad!

### custom predicates

A convenient feature of `assertr` is that it makes the construction of custom
predicate functions easy.

In order to make a custom predicate, you only have to specify cases where the
predicate should return FALSE. Let's say that a dataset has an ID column
(named `ID`) that we want to check is not an empty string. We can create a
predicate like this:

```{r}
not.empty.p <- function(x) if(x=="") return(FALSE)
```

and apply it like this:

```{r, eval=FALSE, purl = FALSE}
read.csv("another-dataset.csv") %>%
  assert(not.empty.p, ID) %>%
  ...
```

Let's say that the ID column is always a 7-digit number. We can confirm that
all the IDs are 7-digits by defining the following predicate:

```{r}
seven.digit.p <- function(x) nchar(x)==7
```

A powerful consequence of this easy creation of predicates is that the
`assert` function lends itself to use with lambda predicates (unnamed
predicates that are only used once). The check above might be better written as

```{r, eval=FALSE, purl = FALSE}
read.csv("another-dataset.csv") %>%
  assert(function(x) nchar(x)==7, ID) %>%
  ...
```

Neat-o!


### enter `insist` and predicate 'generators'

Very often, there is a need to dynamically determine the predicate function
to be used based on the vector being checked.

For example, to check to see if every element of a vector is within _n_
standard deviations of the mean, you need to create a `within_bounds`
predicate _after_ dynamically determining the bounds by reading and computing
on the vector itself.

To this end, the `assert` function is no good; it just applies a raw predicate
to a vector. We need a function like `assert` that will apply predicate
_generators_ to vectors, return predicates, and _then_ perform `assert`-like
functionality by checking each element of the vectors with its respective custom
predicate. This is precisely what `insist` does.

This is all much simpler than it may sound. Hopefully, the examples will clear
up any confusion.

The primary use case for `insist` is in conjunction with the `within_n_sds`
predicate generator.

Suppose we wanted to check that every `mpg` value in the `mtcars` data set was
within 3 standard deviations of the mean before finding the average miles
per gallon for each number of engine cylinders. We could write something
like this:

```{r purl = FALSE}

mtcars %>%
  insist(within_n_sds(3), mpg) %>%
  group_by(cyl) %>%
  summarise(avg.mpg=mean(mpg))
```

Notice what happens when we drop that z-score to 2 stardard deviations
from the mean

```{r error=TRUE,  purl = FALSE}
mtcars %>%
  insist(within_n_sds(2), mpg) %>%
  group_by(cyl) %>%
  summarise(avg.mpg=mean(mpg))
```

Execution of the pipeline was halted. But now we know exactly which data point
(and column) violated the predicate that `within_n_sds(3)(mtcars$mpg)`
returned.

Now that's an efficient car!

After the predicate generator, `insist` takes an arbitrary number of columns
just like `assert` using the syntax of `dplyr`'s `select` function. If you
wanted to check that everything in mtcars is within 10 standard deviations
of the mean (of each column vector), you can do so like this:

```{r purl = FALSE}
mtcars %>%
  insist(within_n_sds(10), mpg:carb) %>%
  group_by(cyl) %>%
  summarise(avg.mpg=mean(mpg))
```

Aces!


### combining chains of assertions

Let's say that as part of an automated pipeline that grabs mtcars from an
untrusted source and finds the average miles per gallon for each number of
engine cylinders, we want to perform the following checks...

- that the dataset contains more than 10 observations
- that the column for 'miles per gallon' (mpg) is a positive number
- that the column for 'miles per gallon' (mpg) does not contain a
datum that is outside 4 standard deviations from its mean, and
- that the am and vs columns (automatic/manual and v/straight engine, 
respectively) contain 0s and 1s only

This could be written thusly:

```{r purl = FALSE}
mtcars %>%
  verify(nrow(mtcars) > 10) %>%
  verify(mpg > 0) %>%
  insist(within_n_sds(4), mpg) %>%
  assert(in_set(0,1), am, vs) %>%
  group_by(cyl) %>%
  summarise(avg.mpg=mean(mpg))
```

Ew, there are four lines of assertions before the real fun starts. We can
make look much better by abstracting out all the assertions:

```{r purl = FALSE}

check_me <- . %>%
  verify(nrow(mtcars) > 10) %>%
  verify(mpg > 0) %>%
  insist(within_n_sds(4), mpg) %>%
  assert(in_set(0,1), am, vs)

mtcars %>%
  check_me %>%
  group_by(cyl) %>%
  summarise(avg.mpg=mean(mpg))
```

Awesome! Now we can add an arbitrary number of assertions, as the need arises,
without touching the real logic.


### advanced: send email reports using custom error functions

One particularly cool application of `assertr` is to use it as a data integrity
checker for frequently updated data sources. A script can download new data as
it becomes available, and then run `assertr` checks on it. This makes assertr
into a sort of "continuous integration" tool (but for data,
not code.)

In an unsupervised "continuous integration" environment, you need a way to
discover that the assertions failed. In CI-as-a-service in the software world, 
failed automated checks often send an email of reporting the maintainer of a
botched build; why not bring that functionality to `assertr`?!

All assertion verbs in `assertr` support a custom error function. By default,
the error function is R's built-in `stop` function, which displays a message
and halts execution. You can specify your own to hijack this behavior and
redirect flow-of-control wherever you want. 

All custom error functions take one argument: the actual error message that
`assertr` assembles based on which assertions fail. In this case, we will
build a function that takes that message and emails it to someone before
halting execution. We will use the `mailR` package to send the mail.

```{r eval=FALSE, purl = FALSE}

library(mailR)

email_me <- function(err_str){
  send.mail(from="assertr@gmail.com", to="YOU@gmail.com",
            subject="error from assertr", body=err_str,
            smtp = list(host.name="aspmx.l.google.com", port=25),
            authenticate = FALSE, send=TRUE)
  stop(err_str, call.=FALSE)
}

questionable_mtcars %>%
  verify(nrow(.) > 10, error_fun=email_me) %>%
  verify(mpg > 0, error_fun=email_me) %>%
  insist(within_n_sds(4), mpg, error_fun=email_me) %>%
  assert(in_set(0,1), am, vs, error_fun=email_me)
  # ...
```

(this particular `send.mail` formulation will only for gmail receptients;
see the `mailR` documentation for more information)

Now you'll get notified of any failed assertions via email. Groovy!


### advanced: creating your own predicate generators for `insist`

`assertr` is build with robustness, correctness, and extensibility in mind.
Just like `assertr` makes it easy to create your own custom predicates, so
too does this package make it easy to create your own custom predicate
generators.

Okay... so its, perhaps, not _easy_ because predicate generators by nature
are functions that return functions. But it's possible!

Let's say you wanted to create a predicate generator that checks if all
elements of a vector are within 3 times the vector's interquartile range from
the median. We need to create a function that looks like this

```{r purl = FALSE}
within_3_iqrs <- function(a_vector){
  the_median <- median(a_vector)
  the_iqr <- IQR(a_vector)
  within_bounds((the_median-the_iqr*3), (the_median+the_iqr*3))
}
```

Now, we can use it on `mpg` from `mtcars` like so:

```{r purl = FALSE}
mtcars %>%
  insist(within_3_iqrs, mpg) %>%
  group_by(cyl) %>%
  summarise(avg.mpg=mean(mpg))
```

There are two problems with this, though...

1. We may want to abstract this so that we can supply an arbitrary number
of IQRs to create the bounds with
2. We lose the ability to choose what optional arguments (if any) that we
give to the returned `within_bounds` predicate.

Now we have to write a function that returns a function that returns
a function...

```{r purl = FALSE}
within_n_iqrs <- function(n, ...){
  function(a_vector){
    the_median <- median(a_vector)
    the_iqr <- IQR(a_vector)
    within_bounds((the_median-the_iqr*n), (the_median+the_iqr*n), ...)
  }
}
```

Much better! Now, if we want to check that every `mpg` from `mtcars` is
within 5 IQRs of the median and *not allow NA values* we can do so like this:

```{r purl = FALSE}
mtcars %>%
  insist(within_n_iqrs(5), mpg) %>%
  group_by(cyl) %>%
  summarise(avg.mpg=mean(mpg))
```

Super!
